# Read in count data
```{r}
library(DESeq2)
library(tidyverse)
library(org.Mm.eg.db)
library(ComplexHeatmap)
library(clusterProfiler)
load("/projects/dylan/221117_overfeeding/y_1_3.rda")
```
# Run DESeq2
```{r fig.width=5, fig.height=3}
mat <- y_1_3$counts
meta <- y_1_3$samples
table(meta$group)
dds <- DESeqDataSetFromMatrix(countData = mat, colData = meta, design = ~ 0+group)

# filter lowly expressed genes
keep <- rowSums(counts(dds) >= 50) >= 5
dds <- dds[keep,]
dds <- DESeq(dds)
vsd <- vst(dds, blind = F)
plotPCA(vsd, intgroup = "group")
```
# Generate DESeq2 results for pairwise comparisons
```{r}
# compare OF and ctrl and D14
res_14 <- 
  results(dds, contrast = c("group", "Overfeeding_day_14", "Control_day_14")) %>% 
  data.frame() %>% 
  rownames_to_column("ENSEMBL") %>% 
  inner_join(
    AnnotationDbi::select(x = org.Mm.eg.db, keys = .$ENSEMBL, columns = "SYMBOL", keytype = "ENSEMBL")
  ) %>% 
  arrange(padj) %>% 
  mutate(comparison = "Day14")

# compare OF and ctrl and D17
res_17 <- 
  results(dds, contrast = c("group", "Overfeeding_day_17", "Control_day_17")) %>% 
  data.frame() %>% 
  rownames_to_column("ENSEMBL") %>% 
  inner_join(
    AnnotationDbi::select(x = org.Mm.eg.db, keys = .$ENSEMBL, columns = "SYMBOL", keytype = "ENSEMBL")
  ) %>% 
  arrange(padj) %>% 
  mutate(comparison = "Day17")

# bind_rows(res_14, res_17) %>% 
#   write_csv("degenes_overfeeding.csv")
```
# Generate heatmap from top DEGs
```{r}
genes <- assay(vsd) %>% data.frame() %>% 
  rownames_to_column("ENSEMBL") %>% 
  inner_join(
    AnnotationDbi::select(x = org.Mm.eg.db, keys = .$ENSEMBL, columns = "SYMBOL", keytype = "ENSEMBL")
  ) %>% 
  filter(SYMBOL %in% c("Igfbp2", "Npy", "Agrp","Fabp7", "Ptgds", 
                       "Stat3","Brwd3","Slc5a7","Chrm2", "Prrg1", 
                       "Hmgcs2", "Riok2", "Myh6", "Gpr89","Ppib",
                       "Bc1","Gkn3", "Tmsb10", "Lars2", "Rps29")) %>% 
  column_to_rownames("SYMBOL") %>% 
  dplyr::select(-c(ENSEMBL)) 

genes <- t(scale(t(genes)))
genes[genes>2] <- 2
genes[genes< -2] <- -2

ha = HeatmapAnnotation(
    Diet = vsd$Condition1[order(vsd$group)], 
    Day = vsd$Condition2[order(vsd$group)],
    col = list(Diet = c("Overfeeding" = "#777acd", "Control" = "#7aa456"), Day = c("day 14" = "#c65999", "day 17" = "#c96d44"))
)

hmap <- ComplexHeatmap::Heatmap(genes[c("Igfbp2", "Npy", "Agrp","Fabp7", "Ptgds", "Stat3","Brwd3","Slc5a7","Chrm2", "Prrg1", "Hmgcs2", "Riok2", "Myh6", "Gpr89","Ppib", "Bc1","Gkn3", 
                                "Tmsb10", "Lars2", "Rps29"),order(vsd$group)], rect_gp = gpar(col = "black"), show_column_names = F, row_title = NULL, 
                                heatmap_legend_param = list(legend_direction = "horizontal", title = "z-score"),
                       cluster_columns = F, cluster_rows = F,column_split = vsd$group[order(vsd$group)], row_split = rep(1:4, each=5), column_title = NULL, top_annotation = ha) 

#svg("heatmap.svg", width = 8, h=5)
draw(hmap, heatmap_legend_side="bottom")
#dev.off()
```

# Try lfc shrinkage
```{r}
dds$group <- relevel(dds$group, ref = "Control_day_14") 
# re-generate model matrix for specified comparison
mm <- model.matrix(~ group, data = colData(dds))
# run nBinomWaldTest with new matrix
ddsshrink <- nbinomWaldTest(dds, modelMatrix = mm)
resultsNames(ddsshrink)

lfc14shrink <- lfcShrink(ddsshrink, coef=2) %>%  data.frame() %>%  
  arrange(-abs(log2FoldChange)*-log10(padj)) %>% 
  rownames_to_column("ENSEMBL") %>% 
  inner_join(
    AnnotationDbi::select(x = org.Mm.eg.db, keys = .$ENSEMBL, columns = "SYMBOL", keytype = "ENSEMBL")
  ) %>% 
  filter(!is.na(padj), !is.na(SYMBOL)) 

lfc14shrink %>% arrange(-log2FoldChange) 
```
# shrink lfc
```{r}
dds$group <- relevel(dds$group, ref = "Control_day_17") 
# re-generate model matrix for specified comparison
mm <- model.matrix(~ 0+group, data = colData(dds))
# run nBinomWaldTest with new matrix
ddsshrink <- nbinomWaldTest(dds, modelMatrix = mm)
resultsNames(dds)
# find log fold change for genes
lfc17shrink <- lfcShrink(ddsshrink, coef=4) %>%  data.frame() %>%  
  arrange(-abs(log2FoldChange)*-log10(padj)) %>% 
  rownames_to_column("ENSEMBL") %>% 
  inner_join(
    AnnotationDbi::select(x = org.Mm.eg.db, keys = .$ENSEMBL, columns = "SYMBOL", keytype = "ENSEMBL")
  ) %>% 
  filter(!is.na(padj), !is.na(SYMBOL)) 

lfc17shrink %>% arrange(-log2FoldChange)
```
# Correlation plot of lfc at both timepoints
```{r fig.width=8, fig.height=6}
test <- inner_join(res_14, res_17, by = c("ENSEMBL")) %>% filter(padj.x < .1 | padj.y < .1)
cor(test$log2FoldChange.x, test$log2FoldChange.y, use="complete.obs", method = "pearson")^2

#svg("corgenes.svg", width = 8, h=5)

test %>% 
  ggplot() +
  aes(log2FoldChange.x, log2FoldChange.y, fill = case_when(pvalue.x<0.05&pvalue.y>0.05 ~ "D14 only",
                                                            pvalue.x>0.05&pvalue.y<0.05 ~ "D17 only",
                                                            pvalue.x<0.05&pvalue.y<0.05 ~ "Both")) +
  geom_point(alpha=0.75, shape=21, stroke=0.1) +
  geom_hline(yintercept = 0, linetype=2) +
  geom_vline(xintercept = 0, linetype=2) +
  geom_smooth(method = "lm", se=F, aes(group=1), color="black", size=0.5) +
  #ggrepel::geom_text_repel(aes(label = ifelse(SYMBOL.x %in% c("Agrp","Npy","Plvap"), SYMBOL.x, NA_character_)), color="black") +
  labs(x = "D14 log2FC (Overfed/Control)", y= "D17 log2FC (Overfed/Control)", fill = "DE") + scale_fill_manual(values = c("#9e6ebd","#7aa457","#cb6751")) +
  guides(fill = guide_legend(override.aes = list(linetype = 0, size=4))) +
  #lims(x=c(-2,2), y=c(-2,2)) +
  coord_fixed()
#dev.off()

#ggsave("correlated_genes.svg", h=4, w=6)
```

# Volcano Plot
```{r}
v14 <- 
  res_14 %>% 
  filter(abs(log2FoldChange)<2.5) %>% 
  ggplot() +
  aes(log2FoldChange, -log10(pvalue)) + 
  geom_point(aes(fill = I(ifelse(padj<0.05, "red","grey70"))), shape=21, size=2, alpha=0.5) +
  theme_classic() + 
  ggrepel::geom_text_repel(aes(label = ifelse(padj<0.005 & abs(log2FoldChange)>0.5, SYMBOL, NA_character_))) +
  ggtitle("Day 14") + 
  theme(plot.title = element_text(hjust=0.5, face = "bold")) +
  lims(y=c(0,20))

v17 <- 
  res_17 %>% 
  filter(abs(log2FoldChange)<2.5) %>% 
  ggplot() +
  aes(log2FoldChange, -log10(pvalue)) + 
  geom_point(aes(fill = I(ifelse(padj<0.05, "red","grey70"))), shape=21, size=2, alpha=0.5) +
  theme_classic() + 
  ggrepel::geom_text_repel(aes(label = ifelse(grepl("Npy|Rp|Lars|Hmg", SYMBOL), 
                                              SYMBOL, NA_character_)), max.overlaps = 50) +
  ggtitle("Day 17") + 
  theme(plot.title = element_text(hjust=0.5, face = "bold")) +
  lims(y=c(0,20))
```

```{r fig.width=8, fig.height=4}
v14+v17
```
# plot expression of DEGs at both timepoints
```{r}
gene_plots <- 
  res_14 %>% 
  filter(log2FoldChange<0, !is.na(SYMBOL), padj < 0.05) %>% 
  dplyr::slice(1:4) %>% 
  bind_rows(
    res_14 %>% 
      filter(log2FoldChange>0, !is.na(SYMBOL), padj < 0.05) %>%
      dplyr::slice(1:4)
  ) %>% 
  pull(ENSEMBL) %>% 
  purrr::map(., function(x) {
    gene <- res_14 %>% filter(ENSEMBL=={{x}}) %>%  pull(SYMBOL)
    plot <-
      coef(ddsshrink)[x,] %>% 
      enframe() %>% 
      inner_join(
        coef(ddsshrink, SE=T)[x,] %>% 
          enframe(value = "SE") %>% 
            mutate(name = gsub("SE_","", name))
      ) %>% 
      mutate(day = gsub(".*_","", name),
             diet = gsub("_.*","", gsub("group","", name))) %>% 
      ggplot() +
      aes(day, value, color=diet) +
      geom_point(position=position_dodge(width=0.5)) +
      geom_errorbar(aes(ymin = value-SE, ymax=value+SE, width=0.1),position=position_dodge(width=0.5)) +
      theme_bw() +
      ggtitle(gene) +
      theme(plot.title = element_text(hjust=0.5, face="bold")) + 
      labs(color="Condition", x="Timepoint", y="Value") +
      theme(legend.position = "none")
  })

cowplot::plot_grid(plotlist = gene_plots, nrow = 2)
```
# plot expression of DEGs at both timepoints
```{r fig.width=15, fig.height=5}
gene_plots_17 <- 
  res_17 %>% 
  filter(log2FoldChange<0, padj<0.05, !is.na(SYMBOL)) %>% 
  dplyr::slice(1:4) %>% 
  bind_rows(
    res_17 %>% 
      filter(log2FoldChange>0, padj<0.05, !is.na(SYMBOL)) %>%
      dplyr::slice(1:4)
  ) %>% 
  pull(ENSEMBL) %>% 
  purrr::map(., function(x) {
    gene <- res_17 %>% filter(ENSEMBL=={{x}}) %>%  pull(SYMBOL)
    plot <-
      coef(ddsshrink)[x,] %>% 
      enframe() %>% 
      inner_join(
        coef(ddsshrink, SE=T)[x,] %>% 
          enframe(value = "SE") %>% 
            mutate(name = gsub("SE_","", name))
      ) %>% 
      mutate(day = gsub(".*_","", name),
             diet = gsub("_.*","", gsub("group","", name))) %>% 
      ggplot() +
      aes(day, value, color=diet) +
      geom_point(position=position_dodge(width=0.5)) +
      geom_errorbar(aes(ymin = value-SE, ymax=value+SE, width=0.1),position=position_dodge(width=0.5)) +
      theme_bw() +
      ggtitle(gene) +
      theme(plot.title = element_text(hjust=0.5, face="bold")) + 
      labs(color="Condition", x="Timepoint", y="Value") +
      theme(legend.position = "none")
  })

cowplot::plot_grid(plotlist = gene_plots_17, nrow = 2)
```
# Run GSEA to identify enriched pathways at D14
```{r fig.width=12, fig.height=10}
genelist14 <- res_14$stat[!is.na(res_14$padj)]
names(genelist14) <- res_14$ENSEMBL[!is.na(res_14$padj)]
genelist14 <- genelist14[!is.na(genelist14)]
genelist14 <- sort(genelist14, decreasing = T)
egobp <- gseGO(geneList     = genelist14,
              OrgDb        = org.Mm.eg.db,
              ont          = "ALL",
              keyType      = "ENSEMBL", 
              minGSSize    = 10,
              maxGSSize    = 300,
              pvalueCutoff = 0.05,
              verbose      = FALSE)

res <- 
  egobp@result %>% 
  mutate(dir = sign(NES)) 

calc_gosim <- function(x) {
  gomat <- x %>%  filter(p.adjust < 0.05)
  ont = unique(x$ONTOLOGY)
  simMatrix <- rrvgo::calculateSimMatrix(gomat$ID,orgdb="org.Mm.eg.db",ont=ont,method="Rel")
  scores <- setNames(-log10(gomat$p.adjust), gomat$ID)
  reducedTerms <- rrvgo::reduceSimMatrix(simMatrix, scores, threshold=0.9,orgdb="org.Mm.eg.db")
}

redterms <- map(split(res, interaction(res$ONTOLOGY, res$dir)), ~calc_gosim(.x))
bind_rows(redterms, .id="comp") %>% 
  group_by(comp, cluster) %>% 
  dplyr::slice_max(n = 1, order_by = score, with_ties = F) %>% 
  write_csv("goterms_reduced_14.csv")
```
# Run GSEA to identify enriched pathways at D14+3
```{r}
genelist17 <- res_17$stat
names(genelist17) <- res_17$ENSEMBL[!is.na(res_17$padj)]
genelist17 <- genelist17[!is.na(genelist17)]
genelist17 <- sort(genelist17, decreasing = T)
egobp17 <- gseGO(geneList     = genelist17,
              OrgDb        = org.Mm.eg.db,
              ont          = "ALL",
              keyType      = "ENSEMBL", 
              minGSSize    = 10,
              maxGSSize    = 300,
              pvalueCutoff = 0.05,
              verbose      = FALSE)

res17 <- 
  egobp17@result %>% 
  mutate(dir = sign(NES)) 

calc_gosim <- function(x) {
  gomat <- x %>%  filter(p.adjust < 0.05)
  ont = unique(x$ONTOLOGY)
  simMatrix <- rrvgo::calculateSimMatrix(gomat$ID,orgdb="org.Mm.eg.db",ont=ont,method="Rel")
  scores <- setNames(-log10(gomat$p.adjust), gomat$ID)
  reducedTerms <- rrvgo::reduceSimMatrix(simMatrix, scores, threshold=0.9,orgdb="org.Mm.eg.db")
}

redterms <- map(split(res17, interaction(res17$ONTOLOGY, res17$dir)), ~calc_gosim(.x))
bind_rows(redterms, .id="comp") %>% 
  group_by(comp, cluster) %>% 
  dplyr::slice_max(n = 1, order_by = score, with_ties = F) %>% 
  write_csv("goterms_reduced_17.csv")
```


